---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Data loaded using read.csv after which it is filtered by removing NA rows. hist is used to plot a histogram
```{r load, echo = TRUE}
data <- read.csv("activity.csv",colClasses = "character")
stepsVector  <- data$steps
filteredData <- data[!is.na(stepsVector),]

filteredData[, 1] <- as.numeric(filteredData[, 1])
filteredData[, 2] <- as.Date(filteredData[, 2],"%Y-%m-%d")

hist(filteredData$steps,xlab = "Steps", ylab = "Frequency", main = "Data Histogram")
```



## What is mean total number of steps taken per day?
Mean steps calculated using mean function and median calculated using median function
```{r daily1, echo = TRUE}
meanSteps <- mean(filteredData$steps)
medianSteps <- median(filteredData$steps)
print(meanSteps)
print(medianSteps)
```



## What is the average daily activity pattern?
for this question, we will need to write a helper function to compute daily steps. The code is as follows
```{r daily2, echo = TRUE}
calcTotalForDay <- function (reqDay,data) {
    dateVector <- data[, 2]
    dayData <- data[dateVector == reqDay,]
    
    d <- c(sum(dayData$steps),reqDay)
    
    d
    
}
```

Once the helper function is done, we can compute daily steps with the following code
```{r daily3, echo = TRUE}
days <- unique(sort(filteredData[,2]))
dailyData <- sapply(days,calcTotalForDay,data = filteredData)

dailyData <- data.frame(t(dailyData))

cnames <- c("steps", "day")
colnames(dailyData) <- cnames

dailyData[, 1] <- as.numeric(as.character(dailyData[, 1]))
```

And then compute daily mean and median using
```{r daily4, echo = TRUE}
hist(dailyData$steps,xlab = "Daily Steps", ylab = "Frequency", main = "Daily Histogram")
totalDailySteps  <- dailyData$steps
meanDailySteps   <- mean(totalDailySteps)
medianDailySteps <- median(totalDailySteps)
```

The mean daily steps is 10766
The median daily steps is 10765

A histogram is shown below

![Histogram of daily steps] (./figures/daily_steps_histogram.png.png)


To compute the most active 5 minute interval, we need another helper function

```{r daily5, echo = TRUE}


fiveMinsIntervalForDay <- function (reqDay,data) {
    dateVector <- data[, 2]
    dayData <- data[dateVector == reqDay,]
    
    d <- dayData$steps
    
    d
    
}
```


With this helper function, we can proceed. First we compute a table with all 5 minute interval accross all days and then name the columns

```{r daily6, echo = TRUE}
dailyPer5MinsData <- sapply(days,fiveMinsIntervalForDay,data = filteredData)
dailyPer5MinsData <- data.frame(t(dailyPer5MinsData))
cnames <- (c(1:length(dailyPer5MinsData))-1)*5
cnames <- c(1:length(dailyPer5MinsData))-1
cnames <- as.character(cnames)
colnames(dailyPer5MinsData) <- cnames

per5MinuteSum <- colSums (dailyPer5MinsData)
print(dim(dailyPer5MinsData))
print(length(per5MinSum))

#plot
plot(per5MinuteSum, type="l", col="blue",xlab = "5 Minute Interval", ylab = "Frequency", main = "Daily Steps per 5 Minute")

# which 5 min interval is max
print(max(per5MinuteSum))
mostActive5Minute <- per5MinuteSum[per5MinuteSum == max(per5MinuteSum)]
print(mostActive5Minute)
```
The most active 5 minute interval is the 103rd  5 minute reading and its value is 10927


![ Time series plot of the 5-minute interval and the average number of steps taken] (./figures/per_5minute_sum.png)

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

